---
title: "Estimating parameters of (non-)stationary GEV models and their estimation uncertainty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{confint_slbm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
library(slbm)
library(dplyr)
library(tidyr)
library(ggplot2)
```

This vignette describes the (non-)stationary univariate that can be fitted based on 
sliding block maxima with the `slbm` package. 


## Preliminaries and Model assumptions

The univariate models treated here assume that every block maximum
\[ M_{r,i}^{(t)} = \max\{ X_i^{(t)}, \ldots, X_r^{(t)}, X_1^{(t+1)}, \ldots, X_{i-1}^{(t+1)} \}, \
i = 1, \ldots, r, \ t = 1, \ldots, m,\]
that is computed over a block containing $r$ of the observations $(X_j^{(t)})_{j,t}$ and that starts in year $t$ satisfies 

\[   M_{r,i}^{(t)} \sim \mathrm{GEV}(\mu(t), \sigma(t), \gamma), \]

where $\mathrm{GEV}(\mu(t), \sigma(t), \gamma)$ has distribution function 

\[ F_{(\mu(t), \sigma(t), \gamma)}(x) = \exp\left( - \left(  1 + \gamma \frac{ x - \mu(t)}{\sigma(t)} \right)_+^{-\frac1\gamma} \right). \]
The models for the location and scale parameter are

* stationary model: 

\[ \mu(t) \equiv \mu, \ \sigma(t) \equiv \sigma.   \]

* shift model:

\[ \mu(t) =  \mu + \alpha \mathrm{GMST}(t), \ \sigma(t) \equiv \sigma.   \]

* scale model:

\[ \mu(t) =  \mu \exp\left( \frac{\alpha}{\mu} \mathrm{GMST}(t) \right), \
   \sigma(t) = \sigma \exp\left( \frac{\alpha}{\mu} \mathrm{GMST}(t) \right).   \]
   

The usual definition of a return level is not compatible with the non-stationary
models. One has to choose a *reference climate* for which one wants to estimate the 
Return Level. For example, an estimate for the 100-year RL in the climate of 1970 is given through
\[ F^{-1}_{(\hat\mu(1970), \hat\sigma(1970), \gamma)}(1-1/100), \]
where $\hat\mu(1970)$ and $\hat\sigma(1970)$ are estimates for the location and scale parameter
of the chosen distribution in year 1970. 

The function `fit_gev_univ` can fit these models based on sliding block maxima.
To obtain confidence intervals for the estimated parameters or for estimated return levels, 
one can employ the function `ci_student_boot_sl`. 
As the name suggests, it performs a studentised bootstrap procedure based on sliding block maxima. To obtain a bootstrap sample, a block bootstrap is performed. For the block-bootstrap-parameter $K$ (which has to be chosen by you), this works as follows:

1. The daily 
    observations are grouped into disjoint blocks consisting of the observations from $K$            seasons, yielding $\lceil m/K \rceil$ larger blocks. 
2. Within these larger disjoint blocks, sliding block maxima are computed. 
3. Based on this *full* sample, the parameters (or return levels), and their estimation variances, are estimated.
4. The indices of the larger blocks that will make up the bootstrap sample are chosen by sampling randomly with replacement from the indices $1, \ldots, \lceil m/K \rceil$. 
5. The bootstrap sample is assembled by concatening all block maxima from the larger blocks that were chosen in step 4. 
6. The model is fitted on the bootstrap sample (and, if required, return levels are estimated).
Further, estimates of the variances are computed.
7. For each parameter (or return level) that is to be estimated, $\alpha/2$ and $1-\alpha/2$ quantiles of the studentised bootstrap estimates are computed, where $\alpha$ denotes the confidence level.
8. For each parameter (or return level) that is to be estimated, the quantiles from step 7 
are multiplied by the standard deviations obtained in step 3, and then substracted from the 
estimates of step 3, resulting in the confidence intervals
\[( \hat\vartheta - \widehat{\mathrm{se}}(\hat\vartheta) t^*_{1-\alpha/2}, \hat\vartheta - \widehat{\mathrm{se}}(\hat\vartheta) t^*_{\alpha/2}), \]
where here $\vartheta$ denotes one of the parameters or RLs that are to be estimated. 



## Application on real data

In attribution studies, the shift model is often used for seasonal maxima of 
daily maximum temperatures, while the scale model is usually applied when analysing 
seasonal maxima of daily cumulated precipitation amounts. 

Therefore we will fit the two non-stationary models to some example data sets of the mentioned
types in the following. 

### The shift model

We start with the shift model. The `DailyTempNRW` data is a list with list components `TimeSeries` which contains measurements of daily maximum temperatures, and `SpatialCov` which contains
station metadata. The latter will not be needed during this analysis, because we want to 
fit a univariate model. Therefore we restrict the analysis to the data from the Station `Aachen` and suppose we want to fit a model for summer's daily maximum temperatures.

The `GMST` data contains values of the global mean surface temperature as well as a smoothed version of it. 

We start by loading the data and extracting the relevant observations that were made at Aachen.

```{r}
data("GMST")
data("DailyTempNRW")

# extract observations from Aachen along with the Date of observation. 
dailyAach <- DailyTempNRW$TimeSeries %>% select(Datetime, Aachen)  
# plot the data
ggplot(dailyAach, aes( x = Datetime, y = Aachen)) + geom_line()
#  in the plot we see that from 2011 onwards there is no data, 
# so we choose only years up to 2010:
dailyAach <-  dailyAach %>% filter(lubridate::year(Datetime) <= 2010)
anyNA(dailyAach) # no NA's left
```
Also, we need to extract the matching covariate values from the `GMST` data. 

```{r}
# first and last year with observations:
fl_year <- c(min(lubridate::year(dailyAach$Datetime)), max(lubridate::year(dailyAach$Datetime)))
gmst <- GMST %>% filter(Year >= fl_year[1], Year <= fl_year[2])
```

This gives us `r fl_year[2] - fl_year[1] +1` seasons of observations, and the seasonlength, i.e. the blocksize is `30 + 31+ 31 = 92`.
We want to compute parameter estimates for the shift model along with their 95\% confidence intervals. Further, we want to compute estimates of the 100-year Return Level for the reference climates of years 1970 and 2010, again with confidence intervals. 

The function `ci_student_boot_sl` does all that for us. A look into the function documentation tells us that it takes the following input: the daily observations as a numeric vector, the blocksize, the 
bootstrap parameter $K$, the number of bootstrap repetitions, the temporal covariate (one value for each day with an observation), the model type (i.e. one of `shift`, `scale` or `stationary`), the return periods for which we want to compute the Return Levels, 
the global mean surface temperatures as observed in the reference years and some other fine tuning parameters. 

As mentioned above, in our example the blocksize is 92, and the reference climates for which 
we want to compute the  are 1970 and 2010. Further, we choose the block-bootstrap 
parameter as $K = 4$.
```{r}
blcksz <- 92
refgmst <- gmst %>% filter(Year %in% c(1970, 2010))
res_shift <- ci_student_boot_sl(yy = dailyAach$Aachen, blcksz = blcksz,
                   temp.cov = rep(gmst$smoothedGMST, each = blcksz),
                   B = 200, type = "shift", Tyrl = 100, Kblock = 4,
                   ref_gmst = refgmst$smoothedGMST, estimate_RL = "both"
                   )
res_shift$res_par  # results for parameter estimation
res_shift$res_rl  # results for RL estimation
```
### The scale model 


