---
title: "Spatio-Temporal-GEV"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatio-Temporal-GEV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(slbm)
library(dplyr)
library(tidyr)
```

The function `fit_spgev_sl()` allows to fit a spatio-temporal GEV model to the 
sample of sliding blockmaxima from several stations in space. 
The model is assumed as follows: Let 
\[ X_{t,d}, t = 1, \ldots, T, d = 1, \ldots, D, \]
denote the observations of some variable of intereset on day $t$ at station $d$. 
The sample of sliding block maxima with blocksize $r$ is then given by $(M_{t,d})$
with
\[ M_{t,d} =  \max\{ X_{t,d}, \ldots, X_{t + r -1, d}\}, t = 1, \ldots, T-r+1, d = 1, \ldots, D. \]

Now, the following spatio-temporal GEV model is assumed:

\[ M_{t,d} \sim \mathrm{GEV}(\mu(t,d), \sigma(t,d), \gamma(t,d)),  \]
where
\[ \mu(t,d) = \mathbf{X}_{\mu_s} \mathbf{\mu_s} + \mathbf{X}_{\mu_t}\mathbf{\mu}_t, \]
and similar for $\sigma(t,d)$ and $\gamma(t,d)$, where 
$\mathbf{X}_{\mu_s}$ and $\mathbf{X}_{\mu_t}$ denote spatial and temporal covariates, 
respectively.


Insert further explanation and formula for weighted $\log$-Likelihood.


### Worklflow

For the synthetic dataset contained in the package.

```{r}
data("ExampleData", package = "slbm")
data("GMST", package = "slbm")
```
`GMST` contains values of the global mean surface temperature anomaly, which is 
often considered as THE temporal covariate.  Because the synthetic dataset contains
40
blocks of 90 observations each (corresponding to 40 years of observations from a 
season of length 90), we have to prepare the temporal covariate accordingly.
Further, we reconstruct the same spatial covariates that were used when generating 
the data. 
```{r}
tempcv <- GMST %>% dplyr::filter(Year %in% c(1980:2019))
tempcvsl <- rep(tempcv$smoothedGMST, each = 90)[1:(39*90 +1)]
tempcvsl <- data.frame(GMST = tempcvsl)
set.seed(3)
spatial_cvrt <- data.frame(lat  = seq(0, 8, length = 8),
  lon = runif(8), ele = runif(8))
```

Now we compute (stationwise) the unique combinations of temporal covariate and values of sliding block maxima of blocksize $r$ 
and their respective frequency of occurence, saved to the variable $n$. This is done with the function 
`get_uniq_bm()`.

```{r}
yy <- data.frame(ExampleData) %>%
      tidyr::pivot_longer(1:8, names_to = "Station", values_to = "Obs")
bmuniq <- get_uniq_bm(yy, 90, temp_cvrt = tempcvsl$GMST, looplastblock = FALSE)
bmuniq
bmuniq[1, ]$uniq_data
```

Now we also compute the sample of disjoint block maxima, because we want to compute
starting values based on disjoint block maxima.
```{r}
djbm <- apply(ExampleData, 2, blockmax, r = 90, "disjoint")

```
Now we can estimate the parameters of our model. 
The spatial parts of the linear models for the location and scale parameter are
passed as R formulas via the `loc.sp.form` and `scale.sp.form` arguments, the temporal formulas via 
the `loc.temp.form` and `scale.temp.form` arguments (also as R formulas).

```{r}
fit_spgev_sl(data = bmuniq, loc.sp.form = ~ lon + lat,
             scale.sp.form = ~ lon +lat,
             loc.temp.form = ~ GMST, spat.cov = spatial_cvrt,
             datastart = djbm, st_val_meth = "LeastSqTemp",
             temp.cov = tempcv$smoothedGMST)

```

Let's try different dependencies on covariates. Further, let's not use a link function 
for modelling the scale parameter.
```{r}
fit_spgev_sl(data = bmuniq, loc.sp.form = ~ lon,
             scale.sp.form = ~ lon ,
             loc.temp.form = ~ GMST, 
             spat.cov = spatial_cvrt,
             datastart = djbm, st_val_meth = "LeastSqTemp",
             temp.cov = tempcv$smoothedGMST, 
             scale.link = make.link("identity"))
```






